discussion

### 2018-10-08 man/ms_1_material_methods.Rmd
update

During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014 @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were charecterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)

@@

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations of @Reyes2015 for mountain regions [@Reyes2015].

@@

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05.

@@

### Tree sampling 
Tree sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].

>>>>

During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)

@@

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

@@

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05.

@@

### Tree sampling 
Tree sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].

### 2018-10-08 man/ms_2_results.Rmd
update

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Tab

The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S6). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155). Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.

>>>>

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).

The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155). Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.

### 2018-10-08 man/ms_3_discussion.Rmd
update

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones muestran altos valores de resiliencia. Como muestrasn nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).

>>>>

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones presentan altos valores de resiliencia. Como muestrasn nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).
